---
title: "Case Analysis 1: Olympics"
description: |
  Nations that Produced the Most Talented Athletes
author: Jack Kim
date: September 18, 2024
format: html
execute:
  echo: false
  warning: false
  message: false
---

[TidyTuesday Reference](https://github.com/rfordatascience/tidytuesday/tree/master/data/2024/2024-08-06)

[Original Data Source](https://www.kaggle.com/datasets/heesoo37/120-years-of-olympic-history-athletes-and-results/)

Objective: Usually it is the Olympic athletes that receive the fame. However, at the end of the day they each represent their country with pride. Let's find out which nations are represented the most in the top 10 most successful athlete list (measured by medal count). So, which countries have produced the best talents?

```{r}
tuesdata <- tidytuesdayR::tt_load('2024-08-06')

olympics <- tuesdata$olympics
```

```{r}
library(tidyverse)

top_athletes <- olympics |>
  filter(!is.na(medal)) |>
  group_by(name, team) |>
  summarise(total_medals = n()) |>
  ungroup() |>
  arrange(desc(total_medals)) |>
  slice(1:10)

top_athletes

top_nations <- top_athletes |>
  group_by(team) |>
  summarise(count = n()) |>
  arrange(desc(count))

top_nations
```

```{r}
ggplot(top_nations, aes(x = reorder(team, -count), y = count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Number of Top 10 Athletes by Country",
       x = "Country (Team)",
       y = "Number of Athletes") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


USSR and United States produced the highest number of most successful athletes. Russia (formerly USSR) and the US usually rank top 3 for medal counts. It is actually surprising to find China not represented in the list.

As for most successful athletes, Michael Phelps is by far has won the most medals (28).
