---
title: "Case Analysis 2: Premier League 2021-2022"
description: |
  Referees in Arsenal's Fixtures
author: Jack Kim
date: September 18, 2024
format: html
execute:
  echo: false
  warning: false
  message: false
---

## Referee Decisions for Arsenal in 2021-2022 Season

[TidyTuesday Reference](https://github.com/rfordatascience/tidytuesday/tree/master/data/2023/2023-04-04)

[Original Data Source](https://www.kaggle.com/datasets/evangower/premier-league-match-data)

**Objective:** Which referees gave out the most yellow cards to Arsenal during the 2021-2022 season?

Refereeing decisions play huge outcome in a football match. In fact, fans and analysts alike do pay enormous attention to the designated referee for each fixture. As a lifelong Arsenal fan, I wanted to find out which referees tend to be harsher on Arsenal.

```{r}
library(tidytuesdayR)

tuesdata <- tidytuesdayR::tt_load('2023-04-04')
tuesdata <- tidytuesdayR::tt_load(2023, week = 14)

soccer <- tuesdata$soccer

```

```{r}
library(tidyverse)

arsenal_yellow_by_referee <- soccer |>
  filter(HomeTeam == "Arsenal" | AwayTeam == "Arsenal") |>
  group_by(Referee) |>
  summarise(
    Home_Yellow_Cards = sum(HY[HomeTeam == "Arsenal"], na.rm = TRUE),
    Away_Yellow_Cards = sum(AY[AwayTeam == "Arsenal"], na.rm = TRUE),
    Total_Yellow_Cards = Home_Yellow_Cards + Away_Yellow_Cards
  ) |>
  arrange(desc(Total_Yellow_Cards))

```


### 1. Which referees handed out the most yellow cards?
```{r, fig.height=5, fig.width=6}
library(ggplot2)

ggplot(arsenal_yellow_by_referee, aes(x = reorder(Referee, -Total_Yellow_Cards), y = Total_Yellow_Cards)) +
  geom_bar(stat = "identity", fill = "yellow") +
  geom_text(aes(label = Total_Yellow_Cards), vjust = -0.5, size = 4) +
  labs(title = "Yellow Cards Arsenal Received by Referees (2021-2022 Season)", 
       x = "Referee", 
       y = "Total Yellow Cards") +
  scale_y_continuous(breaks = NULL) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10), 
    axis.text.y = element_blank(),  
    axis.ticks.y = element_blank(), 
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()  
  )
```

The bar chart represents the total number of yellow cards each referee has given to Arsenal. Craig Pawson handed out the most yellow cards to Arsenal, with 9 cards given.


### 2. Did Arsenal receive more yellow cards at home or away?

```{r}
library(ggplot2)
library(dplyr)

# Summarize total yellow cards at home and away
yellow_cards_summary <- arsenal_yellow_by_referee %>%
  summarise(
    Home = sum(Home_Yellow_Cards),
    Away = sum(Away_Yellow_Cards)
  ) %>%
  pivot_longer(cols = everything(), names_to = "Location", values_to = "Total_Yellow_Cards")

# Create the bar chart
ggplot(yellow_cards_summary, aes(x = Location, y = Total_Yellow_Cards, fill = Location)) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(aes(label = Total_Yellow_Cards), vjust = -0.2, size = 5) + # Adjust vertical position of text
  labs(
    title = "Yellow Cards Arsenal Received by Location",
    x = "Location",
    y = "Total Yellow Cards"
  ) +
  theme_minimal() +
  theme(
    plot.margin = margin(10, 10, 20, 10), # Add extra space below and above the plot
    text = element_text(size = 14),
    legend.position = "none"
  ) +
  ylim(0, max(yellow_cards_summary$Total_Yellow_Cards) + 5) # Increase y-axis limit to prevent clipping


```

This study offers insight into if Arsenal experienced home advantage in referee decisions. Many fans assume that referees tend to be more lenient toward home teams. This may manifest the most in referee decisions for making disciplinary actions. The bar chart shows the number of yellow cards Arsenal received depending on match venue.

Interestingly, Arsenal received 29 yellow cards at away matches and 31 in home matches. This goes against common logic. Although not conclusive, home advantage did not work in Arsenal's favor.
